Verification of OpenSSL's SHA3 \keccak\ function using the Software Analysis Workbench

### 2022-01-06 paper/fips202.tex
Minor revision and clarification

This section illustrates with a few examples how the \fips\ specification is captured in \cryptol. 
\cryptol's functional folds and list comprehensions do not directly align with the quantification and iteration in the specification. 
This misalignment made the \cryptol\ specification difficult to read and to manually argue its equivalence relative to the published specification.
The two specifications are realigned through a library of methods in \cryptol\ that hide the functional folds and list comprehensions so that the \cryptol\ specification reads more like the \fips\ specification. The two library methods are the \emph{for-method} and the \emph{while-method} discussed in this section.

@@

\newsavebox{\PiCry}
\begin{lrbox}{\PiCry}
  \begin{lstlisting}[language=Cryptol, numbers=left, escapeinside={;}{;}]
    type STATE_ARR = [5][5][64];\label{line:state};
    pi : STATE_ARR -> STATE_ARR
    pi a = [  [  [a @x @((x + 3*y) % 5) @z;\label{line:pi};
                 | z <- [0..63]];\label{line:list};
              | y <- [0..4]] 
           | x <- [0..4]]
  \end{lstlisting}
\end{lrbox}

@@

The \cryptol\ definition using list comprehensions is given in \figref{fig:pi}(b).
\lineref{line:state}) defines the state as a $5 \times 5$ array of 64-bit words.
\lineref{line:pi} starts the definition of $\pi$. 
Here the comprehensions are nested with the list indices appearing in reverse order to follow the nesting.
\lineref{line:list} gives the domain for $z$ as a list with values from 0 to 63.
The domains for $x$ and $y$ are given similarly.
The domains for $x$, $y$, and $z$ all match the domains defined in the quantification in \figref{fig:pi}(a).
Although the list comprehensions are not unreadable, the \cryptol\ can be improved by creating a for-method to hide some of the details.

@@

\figref{fig:pi}(d) is the rewritten \cryptol\ using the for-method.
It still requires the indices lists for the comprehension, but it has a more obvious correspondence to the original definition in \figref{fig:pi}(a).
The for-method reorders and structures the arguments in a very elegant way.
It hides the list comprehensions to simplify the task of visual inspection for equivalence.

@@

The \emph{rc} function checks the value of the input integer $t$, and returns the bit $0$ if it is a multiple of $255$.
If not, then Line 2 defines $R$ as the 8-bit value $128$, and Line 3 iteratively transforms $R$ a fixed number of times: $t\ \mathbf{mod}\ 255$.
The transform is defined on Line 3a through Line 3f.
Line 3a prepends a $0$ bit to $R$ so that it is now a 9-bit value. The array indexing refers to bit positions from left to right so $R[0]$ is the prepended left-most bit and $R[8]$ is the right-most bit. Each transform on $R$ prepends the 0 (Line 3a), does an exclusive-or with specific bit positions (Line 3b through Line 3e), and then takes the left-most 8 bits for the next round effectively dropping the right-most bit (Line 3f).
It returns the $R[0]$ at the end.

>>>>

This section illustrates with a few examples how the \fips\ specification is captured in \cryptol. 
\cryptol's list comprehensions do not directly align with the quantification and iteration in the specification. 
This misalignment makes the \cryptol\ specification more difficult to read and to manually argue its equivalence relative to the published specification.
The two specifications are better aligned through a library of two methods in \cryptol\ that hide the list comprehensions so that the \cryptol\ specification reads more like the \fips\ specification. The two library methods are the \emph{for-method} and the \emph{while-method} discussed in this section.

@@

\newsavebox{\PiCry}
\begin{lrbox}{\PiCry}
  \begin{lstlisting}[language=Cryptol, numbers=left, escapeinside={;}{;}]
    type STATE_ARR = [5][5][64];\label{line:pistate};
    pi : STATE_ARR -> STATE_ARR
    pi a = [  [  [a @x @((x + 3*y) % 5) @z;\label{line:pi};
                 | z <- [0..63]];\label{line:list};
              | y <- [0..4]] 
           | x <- [0..4]]
  \end{lstlisting}
\end{lrbox}

@@

The \cryptol\ definition using list comprehensions is given in \figref{fig:pi}(b).
\lineref{line:pistate} defines the state as a $5 \times 5$ array of 64-bit words.
\lineref{line:pi} starts the definition of $\pi$. 
Here the comprehensions are nested with the list indices appearing in reverse order to follow the nesting.
\lineref{line:list} gives the domain for $z$ as a list with values from 0 to 63.
The domains for $x$ and $y$ are given similarly.
The domains for $x$, $y$, and $z$ all match the domains defined in the quantification in \figref{fig:pi}(a).
Although the list comprehensions are not unreadable, the \cryptol\ can be improved by creating a for-method to hide some of the details.

@@

\figref{fig:pi}(d) is the rewritten \cryptol\ using the for-method.
It still requires the indices lists for the list comprehensions, but it has a more obvious correspondence to the original definition in \figref{fig:pi}(a).
The for-method reorders and structures the arguments in a very elegant way.
It hides the list comprehensions to simplify the task of visual inspection for equivalence.

@@

The \emph{rc} function checks the value of the input integer $t$, and returns the bit $0$ if it is a multiple of $255$.
If not, then Line 2 defines $R$ as the 8-bit value $128$, and Line 3 iteratively transforms $R$ a fixed number of times: $t\ \mathbf{mod}\ 255$.
The transform is defined on Line 3a through Line 3f.
Line 3a prepends a $0$ bit to $R$ so that it is now a 9-bit value. The array indexing refers to bit positions from left to right so $R[0]$ is the prepended left-most bit and $R[8]$ is the right-most bit. Each transform on $R$ prepends the 0 (Line 3a), does an exclusive-or with specific bit positions (Line 3b through Line 3e), and then takes the left-most 8 bits for the next round effectively dropping the right-most bit (Line 3f).
It returns the $R[0]$ bit at the end.

### 2022-01-06 paper/openssl.tex
Minor revision and clarification

This section details the two differences between the \openssl\ implementation of \shaThree\ and the \fips\ specification that prevented a direct proof of equivalence between the \cryptol\ and C code.
These differences are expressed in a second \cryptol\ model derived from the \fips\ model discussed in the previous section.
\saw\ proves these two models equivalent for the 256-bit digest size as discussed in \secref{sec:proof}.
This equivalence considers the whole of the sponge construction algorithm proving that the digest is the same from each of them.
That said, and as a reminder, the equivalence between the \keccak\ description in \cryptol\ and the C implementation in \openssl\ holds for any input message and any digest size.
That proof uses the \cryptol\ definition for the model discussed in this section that includes to two differences seen in \openssl, and the overrides used in that proof also come from the model discussed in this section.

\subsection{State Array Structure}

The first difference in the \openssl\ \shaThree\ implementation is in the structure of the state array.
The difference is an artifact of how C maps structured data to memory.
The \fips\ structures the state array as a $5 \times 5$ matrix of 64-bit words with each 64-bit word being a \emph{lane}.
It assumes the memory layout of the data follows a normal cartesian three dimensional coordinate system with $x$ being the horizontal axis, $y$ being the vertical axis, and $z$ being the depth access an a line to get an individual bit from the lane.
For a state $A$, $A[x,y,z]$ accesses $z^\mathrm{th}$ bit from the 64-bit word on the $x^\mathrm{th}$ column and the $y^\mathrm{th}$ row.
That is the mapping to a cartesian coordinate system.

@@

Adding to the complexity is that the C standard does not provide array indexing to get a bit from a value.
For the $A[x,y,z]$ example, there is no array bracket notation to get the $z^\mathrm{th}$ bit in a 64-bit word, so notation in the \fips\ standard such as that seen in the definition of the $rc$ function in \figref{fig:rc}(a) has no direct analogue in C.
As such, the \openssl\ implementation does everything at the level of the lanes, operating on each lane as a 64-bit entity, and never refers to an individual bit in a lane.
Finally, bit ordering in the lanes in the \fips\ standard is just opposite the ordering in C meaning that the direction of shifting in standard is just opposite the direction used in C.

The \cryptol\ for the \fips\ model is rewritten to reflect these memory layout differences.
The $x$ and $y$ indexing is swapped.
All the inner \keccak\ functions are modified to operate on lanes in their entirety.
And the the shifts are reversed.
For reference, \figref{fig:piopenssl} is the rewritten $\pi$ function in \cryptol\ than matches \openssl's implementation (compare to \figref{fig:pi}(d)).
Operating on lanes in faster and more efficient.
That is reflected in the \cryptol\ running times.

>>>>

This section details the key differences between the \openssl\ implementation of \shaThree\ and the \fips\ specification that prevented a direct proof of equivalence between the \cryptol\ and C code.
These differences are expressed in a second \cryptol\ model derived from the \fips\ model discussed in the previous section.
\saw\ proves these two models equivalent for the 256-bit digest size as discussed in \secref{sec:proof}.
This equivalence considers the whole of the sponge construction algorithm proving that the digest is the same from each of them.
That said, and as a reminder, the equivalence between the \keccak\ description in \cryptol\ and the C implementation in \openssl\ holds for any input message and any digest size.
That proof uses the \cryptol\ definition for the model discussed in this section that includes all the differences seen in \openssl, and the overrides used in that proof also come from the \cryptol\ discussed in this section.

\subsection{State Array Structure and Computation}

The first set of differences in the \openssl\ \shaThree\ implementation is in the structure of the state array and how it operates on that state array.
The difference in structure is an artifact of how C maps arrays to memory.
The \fips\ structures the state array as a $5 \times 5$ grid of 64-bit words with each 64-bit word being a \emph{lane}.
It assumes the layout of the data follows a normal cartesian three dimensional coordinate system with $x$ being the horizontal axis, $y$ being the vertical axis, and $z$ being the depth on a lane to access an individual bit.
For a state $A$, $A[x,y,z]$ accesses the $z^\mathrm{th}$ bit from the 64-bit word on the $x^\mathrm{th}$ column and the $y^\mathrm{th}$ row.

@@

Adding to the complexity is that the C standard does not provide array indexing to get a bit from a value.
For the $A[x,y,z]$ example, there is no array bracket notation to get the $z^\mathrm{th}$ bit in a 64-bit word, so notation in the \fips\ standard such as that seen in the definition of the $rc$ function in \figref{fig:rc}(a) has no direct analogue in C.
The consequence is that the \openssl\ implementation does everything at the level of the lanes, operating on each lane as a 64-bit entity, and it never refers to an individual bit in a lane.
Finally, the bit ordering in the lanes in the \fips\ standard is just opposite the ordering in C meaning that the direction of shifting in the standard is opposite the direction used in the C implementation.

The \cryptol\ for the \fips\ model is rewritten to reflect these memory layout and computation differences.
The $x$ and $y$ indexing is swapped.
All the inner \keccak\ functions are modified to operate on lanes in their entirety.
And the the shifts are reversed.
For reference, \figref{fig:piopenssl} is the rewritten $\pi$ function in \cryptol\ than matches \openssl's implementation (compare to \figref{fig:pi}(d)).
Operating on lanes in faster and more efficient.
That is reflected in the \cryptol\ running times.

### 2022-01-06 paper/proof.tex
Revise OpenSSL
Minor revision and clarification

The equivalence between the \fips\ standard and the \cryptol\ specification is argued by visual inspection along with tests over the published test vectors with their expected digests.
\shaThree\ is described algorithmically in the standard making use of universal quantifiers and other looping structures in the definition.
\cryptol\ is functional in that it uses list comprehensions for iteration, and it does not have quantifiers.
Visually certifying a specification of \shaThree\ that uses list comprehensions directly for quantification and looping is difficult at best---the list comprehensions are too complex to read, and they obscure the algorithmic, and mathematic, structure in the published standard.

The syntactic disconnect between the quantification and looping in the published standard and the list comprehensions in the \cryptol\ specification is bridged with a library that hides the list comprehensions with appropriately named functions.
The final specification with the function hiding the list comprehensions has a more direct correspondence to the published standard that is easy to visually certify.
The specification is further validated by tests in \cryptol\ over the published test vectors.
The digests computed from the \cryptol\ specification exactly match those published in the standard on each given input.
The equivalence arrow between the \emph{NIST Standard} box and the \emph{Spec} box represents the visual certification and test vector results.
\secref{sec:fips} details the specification and library.

\subsection{Equivalence between Memory Layouts}

The extracted \sawcore\ from the \openssl\ C implementation of \shaThree\ is too big and complex for \saw\ to reason about directly.
Adding to the complexity of the equivalence proof is that the \openssl\ implementation changes the meaning of the state $S$ in the algorithm so that it no longer directly matches the meaning defined in the published standard.
The change effectively reorders the matrix definition of the state to be more amenable to memory and aspects of the \keccak\ implementation in C.
The change is significant enough to make it extremely difficult to argue manually that the computation implemented in \openssl\ matches that defined in the standard.

@@

\subsection{Overrides and Keccak}

>>>>

The equivalence between the \fips\ standard and the \cryptol\ specification is argued by visual inspection along with tests over the published test vectors with their expected digests.
\shaThree\ is described algorithmically in the standard making use of universal quantifiers and other looping structures in the definition.
\cryptol\ is functional in that it uses list comprehensions for iteration, and it does not have quantifiers.
Visually certifying a specification of \shaThree\ that uses list comprehensions is less direct.
The list comprehensions can be complex, and at times, they obscure the algorithmic, and mathematic, structure in the published standard.

The syntactic disconnect between the quantification and looping in the published standard and the functional nature of the \cryptol\ specification is bridged with a library that hides list comprehensions with appropriately named functions.
The final specification using the library has a more direct correspondence to the published standard that is more directly visually certified.
The specification is further validated by tests in \cryptol\ over the published test vectors.
The digests computed from the \cryptol\ specification exactly match those published in the standard on each given input.
The equivalence arrow between the \emph{NIST Standard} box and the \emph{Spec} box represents the visual certification and test vector results.
\secref{sec:fips} details the specification and library.

\subsection{Memory Layout and Optimized Computation}

The extracted \sawcore\ from the \openssl\ C implementation of \shaThree\ is too big and complex for \saw\ to reason about directly.
Adding to the complexity of the equivalence proof is that the \openssl\ implementation changes the meaning of the state $S$ in the algorithm so that it no longer directly matches the meaning defined in the published standard.
The change effectively reorders the matrix definition of the state to be more amenable to memory, and it changes the different functions in \keccak\ to be more efficient when operating on the state.
The changes are significant enough to make it extremely difficult to argue manually that the computation implemented in \openssl\ matches that defined in the standard.

@@

\subsection{Overrides and \keccak}

### 2022-01-06 paper/saw.tex
Minor revision and clarification

\saw's function, among other things, is to prove input to output equivalence between \sawcore\ models.
It accomplishes such proofs with equivalence preserving rewrites on the \sawcore\ and with automated reasoning.
The rewrites simplify the \sawcore\ models as size correlates with complexity in the automated reasoning. 
The rewrites also try to structurally reduce one model to the other thus avoiding automated reasoning altogether.
When rewriting is not sufficient, the equivalence between two models is reduced to a satisfiability problem and dispatched to external backend solvers.

\saw\ supports hierarchical reasoning with \emph{overrides}.
An override replaces one \sawcore\ model with another in the formal analysis.
As an example, consider a Java method that itself calls another Java method as part of its implementation.
Suppose that \saw\ is able to prove that a \cryptol\ specification for the called method and the Java implementation of the same method are equivalent.
If that is possible, then an override is able to replace the called Java method with the \sawcore\ from the \cryptol\ and as a result avoid any symbolic execution of that method when creating the \sawcore\ model of the Java.

As mentioned previously, \sawcore\ from \cryptol\ is less complex than equivalent models created by symbolic execution from C or Java; thereby overrides reduce the overall complexity of the final extracted \sawcore\ models from C or Java.
As such, proving a property about a complex C function or Java method generally requires the use of overrides from \cryptol\ in order for the equivalence proof to be successful, and that is regardless of how that proof is accomplished.
Overrides improve the likelihood of the rewrites showing equivalence and the likelihood of the backend solver being able to show equivalence if the rewrites fail.
Indeed, the use of overrides are key to the \saw\ proof in this paper showing equivalence between the \shaThree\ specification and the \openssl\ C implementation.

>>>>

\saw's function, among other things, is to prove input to output equivalence between \sawcore\ models.
It accomplishes such proofs with equivalence preserving rewrites on the \sawcore\ and with automated reasoning.
The rewrites simplify the \sawcore\ for a models to reduce the cost of automated reasoning. 
The rewrites also try to structurally reduce one model to the other thus avoiding automated reasoning altogether.
When rewriting is not sufficient, the equivalence between two models is reduced to a satisfiability problem and dispatched to external backend solvers.

\saw\ supports hierarchical reasoning with \emph{overrides}.
An override replaces one \sawcore\ model with another in the formal analysis.
As an example, consider a Java method that itself calls another Java method as part of its implementation.
Suppose that \saw\ is able to prove that a \cryptol\ specification for the called method and the Java implementation of the same method are equivalent.
If that is possible, then an override is able to replace the called Java method with the \sawcore\ from the \cryptol;
thereby, it bypasses the symbolic execution of that called method when creating the \sawcore\ model of the Java by using the \sawcore\ from the \cryptol\ specification directly.

As mentioned previously, \sawcore\ from \cryptol\ is less complex than equivalent models created by symbolic execution from C or Java;
thereby, the overrides reduce the overall complexity of the final extracted \sawcore\ from C or Java.
As such, proving a property about a complex C function or Java method generally requires the use of overrides from \cryptol\ in order for the equivalence proof to be successful, and that is regardless of how that proof is accomplished.
Overrides improve the likelihood of the rewrites showing equivalence and the likelihood of the backend solver being able to show equivalence if the rewrites fail.
Indeed, the use of overrides are key to the \saw\ proof in this paper showing equivalence between the \shaThree\ specification and the \openssl\ C implementation.
